Structured Versus Decorated Cospans

### 2020-12-07 cospans.tex
changed statement of Lemma 3.4, statement and proof of Lemma 3.6

% Structured and decorated cospans
% John Baez, Kenny Courser and Christina Vasilakopoulou
% in LaTeX
% 2020/12/5

@@

These 2-equivalences further restrict to the case when the Grothendieck category $(X,\otimes_\X)$ is specifically cocartesian monoidal itself, with coproducts built up from \cref{eq:explicitstructure2}. In that case, opfibrations $(\X,+)\to(\A,+)$ that strictly preserve coproducts and initial object bijectively correspond to pseudofunctors into the 2-category of cocartesian categories.  For more details, see \cite[Corollary 4.7]{MV} and the related discussion.

@@

\begin{lem}[\textbf{Hermida}] \label{lem:fibrewiselimits}
Suppose $\J$ is a small category and $U\maps\X\to\A$ is an opfibration. If the base $\A$ has $\J$-colimits,
the following are equivalent:
\begin{enumerate}
 \item all fibers have $\J$-colimits, and the reindexing functors preserve them;
 \item the total category $\X$ has $\J$-colimits, and $U$ preserves them.
Moreover in this situation the $\J$-colimits in $\X$ and $\J$ can be chosen so that $U$ strictly preserves them.   %%% CHECK AND MAYBE FIX THIS!!!
\end{enumerate}
\end{lem}

@@

On a highly related note, in what follows we are also interested in the existence of a left adjoint $L$ to an induced monoidal opfibration; the following result provides sufficient conditions for that.

\begin{lem}[\textbf{Gray}] \label{prop:opfibtolari}
  Let $U \maps\X \to \A$ be an opfibration. Then
  $U$ is a right adjoint `left inverse', a.k.a the unit is the identity, if (and only if) its fibers have
  initial objects which are preserved by the
  reindexing functors.
\end{lem}

\begin{proof}
 This is \cite[Proposition~4.4]{Gray}. The left adjoint $L\maps\A\to\X$ maps an object $a$ to the initial object in its fiber $\X_a$, denoted by $\bot_a$. By construction, $U(L(a))=U(\bot_a)=a$. On morphisms $f\maps a\to a'$, $L(f)$ is defined by
 \begin{equation}\label{eq:Lonarrows}
  \bot_a\xrightarrow{\mathrm{Cocart}(f,\bot_a)}f_!(\bot_a)\stackrel{\chi_a}{\cong}\bot_{a'}
 \end{equation}
where $\mathrm{Cocart}(f,\bot_a)$ is the cocartesian lifting of $\bot_a$ along $f$ and $\chi_a$ is the unique isomorphism between the initial objects in the fiber above $a'$ since $f_!$ preserves them.
\end{proof}

@@

Next, we can define a symmetric monoidal double functor
\[      \graysquare \maps F \lCsp \to D \lCsp \]
sending any open Petri net with rates to its corresponding open dynamical system.  This was already defined at the level of categories by Pollard and the first author \cite[Section 7]{BP}, who called it `gray-boxing'.   To boost this result to the double category level we use \cref{thm:functoriality}, taking the square in that theorem to be
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$\Fin\Set$};
\node (B) at (1,0) {$\Cat$};
\node (C) at (0,-1) {$\Fin\Set$};
\node (D) at (1,-1) {$\Cat$.};
\node (E) at (0.5,-0.5) {$\Downarrow \theta$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$F$} (B)
(A) edge node[left]{$1$} (C)
(B) edge node[right]{$1$} (D)
(C) edge node[above]{$D$} (D);
\end{tikzpicture}
\] 
Here $\theta$ is a monoidal natural isomorphism given as follows.  For any finite set $S$, $\theta_S \maps F(S) \to D(S)$ maps any Petri nets with rates 
\[   \xymatrix{ (0,\infty) & T \ar[l]_-r \ar@<-.5ex>[r]_-t \ar@<.5ex>[r]^-s & \N[S] }\]
to an algebraic vector field on $\R^S$, say $v$.   This vector field is defined using a standard prescription taken from chemistry, called `the law of mass action'.   Namely, for any $c \in \R^s$, we set
\[  
v(c) = \sum_{\tau \in T} r(\tau) \, ( t(\tau) - s(\tau) ) c^{s(\tau)} 
\]
where 
\[     c^{s(\tau)} = \prod_{i \in S} {c_i}^{s(\tau)_i}  \]
and we think of $t(\tau), s(\tau) \in \N[S]$ as vectors in $\R^S$.   This formula is explained in the paper with Pollard, where it is also shown that $\theta$ defines a monoidal natural isomorphism between functors to $\Set$ \cite[Theorem 18]{BP}.   As such, it trivially becomes a monoidal natural isomorphism between the functors $F, D \maps \Fin\Set \to \Cat$.

@@

Even though for both \cref{thm:equiv} and the above described setting, the conditions required are only sufficient, they provide certain indication that under a suitable ``condition intersection'' in a sense, we could establish a functorial picture
\begin{displaymath}
\begin{tikzpicture}
\node [align=center,draw] (A) { Lax monoidal pseudofunctors \\ $(\A,+)\to(\Cat,\times)$};
\node[below right=1 and 1.5 of A,align=center,draw] (C)  { Symmetric monoidal \\double categories };
\node[below left=1.05 and .5  of A,draw] (D)  {Special opfibrations};
\node[below left=2 and 1 of A,draw] (E)  {Special ralis};
\node[below=3 of A,align=center,draw] (B)  {Finite colimit preserving \\ functors $A\to\X$};
\node[below=0 of D] {\rotatebox{90}{$\simeq$}};
\path[->,font=\scriptsize]
(A) edge node[above,sloped]{$F\mapsto F\lCsp$}  (C)
(D) edge (A)
(E) edge (B)
(B) edge node[below,sloped]{$L\mapsto {}_{L}\lCsp(\X)$} (C);
\end{tikzpicture}
\end{displaymath}
with an actual natural isomorphism in the middle. This would imply that starting from an appropriate middle ground, employing the structured or the decorated cospan formalism is essentially the same. We leave such considerations for future work.

>>>>

% Structured and decorated cospans
% John Baez, Kenny Courser and Christina Vasilakopoulou
% in LaTeX
% 2020/12/7

@@

These 2-equivalences further restrict to the case when the Grothendieck category $(X,\otimes_\X)$ is specifically cocartesian monoidal itself, with coproducts built up from \cref{eq:explicitstructure2}. In that case, opfibrations $(\X,+) \to (\A,+)$ that strictly preserve coproducts and initial object bijectively correspond to pseudofunctors into the 2-category of cocartesian categories.  For more details, see \cite[Corollary 4.7]{MV} and the related discussion.

@@

\begin{lem}[\textbf{Hermida}] \label{lem:fibrewiselimits}
Suppose $\J$ is a small category and $U\maps\X\to\A$ is an opfibration. If the base $\A$ has $\J$-colimits,
the following are equivalent:
\begin{enumerate}
 \item all fibers have $\J$-colimits, and the reindexing functors preserve them;
 \item the total category $\X$ has $\J$-colimits, and $U$ preserves them.
Moreover in this situation the $\J$-colimits in $\X$ can be chosen so that $U$ strictly preserves them.  
\end{enumerate}
\end{lem}

@@

On a highly related note, in what follows we are also interested in the existence of a left adjoint $L$ to an induced monoidal opfibration.  The following result provides sufficient conditions for that.  Following Gray \cite{Gray}, let us define a `right adjoint left inverse' or  \define{rali} to be a right adjoint in an adjunction for which the unit is the identity.

\begin{lem}[\textbf{Gray}] \label{prop:opfibtolari}
  Let $U \maps\X \to \A$ be an opfibration. Then $U$ is a rali if (and only if) fibers are equipped with initial objects that are preserved up to isomorphism by the reindexing functors.
\end{lem}

\begin{proof}
 This is \cite[Proposition~4.4]{Gray}; we only need the ``if'' direction here.    Suppose each fiber $\X_a$ of the opfibration $U$ is equipped with an initial object $\bot_a$, and these objects are preserved by the reindexing functors.   Define $L\maps\A\to\X$ on objects $a \in \A$ by $L(a) = \bot_a$.   Define $L$ on morphisms as follows: given $f\maps a\to a'$, $L(f)$ is the composite
 \begin{equation}\label{eq:Lonarrows}
  \bot_a\xrightarrow{\mathrm{Cocart}(f,\bot_a)}f_!(\bot_a)\stackrel{\chi_a}{\cong}\bot_{a'}
 \end{equation}
where $\mathrm{Cocart}(f,\bot_a)$ is the cocartesian lifting of $f$ to $\bot_a$  and $\chi_a$ is the unique isomorphism between the initial objects in the fiber above $a'$, since $f_!$ preserves them.   One can make $U$ into a right adjoint of $L$ with unit $a \mapsto U(L(a))$ being the identity, using the fact that $U(L(a)) = U(\bot_a) = a$.
\end{proof}

@@

Next, we can define a symmetric monoidal double functor
\[      \graysquare \maps F \lCsp \to D \lCsp \]
sending any open Petri net with rates to its corresponding open dynamical system.  This was already defined at the level of categories by Pollard and the first author \cite[Section 7]{BP}, who called it `gray-boxing'.   To boost this result to the double category level we use \cref{thm:functoriality}, taking the square in that theorem to be
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$\Fin\Set$};
\node (B) at (1,0) {$\Cat$};
\node (C) at (0,-1) {$\Fin\Set$};
\node (D) at (1,-1) {$\Cat$.};
\node (E) at (0.5,-0.5) {$\Downarrow \theta$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$F$} (B)
(A) edge node[left]{$1$} (C)
(B) edge node[right]{$1$} (D)
(C) edge node[above]{$D$} (D);
\end{tikzpicture}
\] 
Here $\theta$ is a monoidal natural isomorphism given as follows.  For any finite set $S$, $\theta_S \maps F(S) \to D(S)$ maps any Petri nets with rates 
\[   \xymatrix{ (0,\infty) & T \ar[l]_-r \ar@<-.5ex>[r]_-t \ar@<.5ex>[r]^-s & \N[S] }\]
to an algebraic vector field on $\R^S$, say $v$.   This vector field is defined using a standard prescription taken from chemistry, called `the law of mass action'.   Namely, for any $c \in \R^s$, we set
\[  
v(c) = \sum_{\tau \in T} r(\tau) \, ( t(\tau) - s(\tau) ) c^{s(\tau)} 
\]
where 
\[     c^{s(\tau)} = \prod_{i \in S} {c_i}^{s(\tau)_i}  \]
and we think of $t(\tau), s(\tau) \in \N[S]$ as vectors in $\R^S$.   This formula is explained in the paper with Pollard, where it is also shown that $\theta$ defines a monoidal natural isomorphism between functors to $(\Set,\times)$ \cite[Theorem 18]{BP}.   As such, it automatically becomes a monoidal natural isomorphism between the functors $F, D \maps (\Fin\Set,+) \to (\Cat,\times)$.

@@

Even though for both \cref{thm:equiv} and the above result the conditions required are only sufficient, they suggest that with work we could establish a functorial picture
\begin{displaymath}
\begin{tikzpicture}
\node [align=center,draw] (A) { Lax monoidal pseudofunctors \\ $(\A,+)\to(\Cat,\times)$};
\node[below right=1 and 1.5 of A,align=center,draw] (C)  { Symmetric monoidal \\double categories };
\node[below left=1.05 and .5  of A,draw] (D)  {Special opfibrations};
\node[below left=2 and 1 of A,draw] (E)  {Special ralis};
\node[below=3 of A,align=center,draw] (B)  {Finite colimit preserving \\ functors $A\to\X$};
\node[below=0 of D] {\rotatebox{90}{$\simeq$}};
\path[->,font=\scriptsize]
(A) edge node[above,sloped]{$F\mapsto F\lCsp$}  (C)
(D) edge (A)
(E) edge (B)
(B) edge node[below,sloped]{$L\mapsto {}_{L}\lCsp(\X)$} (C);
\end{tikzpicture}
\end{displaymath}
with an actual natural isomorphism in the middle.  
This would imply that starting from an appropriate middle ground, employing the structured or the decorated cospan formalism is essentially the same. We leave such considerations for future work.
